Inference using context-dependent models of mutation

### 2021-10-13 writeups/context-dependence.tex
Finishing pass to the full likelihood

First we quantify and prove the approximation \eqref{eqn:window_approx}.
The basic idea is that the probability is only in error if there is a chain of mutations
that create a dependency between the base of the T-mer and the flanking sequence.
To simplify the argument, first assume that all transition triples
change only a single, central site in the pattern,
taking into account context on each side of maximum length $w$.
(For instance, a trinucleotide model in which the flanking sites on each side
affect the mutation rate of the central site has $w=1$.)
Recall that in our construction of the process we first chose the times and sites
at which potential changes might occur.
Suppose a potential change occurs at site $i-w$, at time $t$,
which therefore might result in a change in state at site $i$.
To resolve this, we need to know the outcome of any other potential changes
that might change the sequence between $i-w$ and $i + w$.
Consider those that might happen only to the right of $i$,
thus extending the sequence on which the change depends to the right.
The number of such potential changes is Poisson with mean $\mu^* w t$,
and will extend the amount of sequence on which the change at site $i$ depends
by a uniform random number of sites between 1 and $w$.
If one occurs, say at time $0 < s < t$,
then any further potential changes that occur between time 0 and time $s$
in the $w$ sites to the right of this older potential change
will again extend the range of dependency.
Putting this together, the amount of sequence to the right of a site
on which a mutation at that site depends
is bounded by the compound Poisson process $\sum_{k=1}^N W_k$,
where $N$ is Poisson with mean $\mu^* w t$
and each $W_k$ is independent and uniform on $\{1, 2, \ldots, w\}$.

Now consider the probability of interest,
\begin{align*}
    \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \} .
\end{align*}
This probability can be written as a sum over all possible arrangements
and resolutions of potential changes.
Our approximation, $p_{n,\ell,r}(x,y)$ is the probability that would be obtained
by ignoring any potential changes that rely on context outside of the $n$-sequence $x$.
So, if $A$ is the event that the final subsequence $y$
depends on seequence outside of $x$ due to potential changes,
then
\begin{align*}
    \left|
        \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
        -
        p_{n,\ell,r}(x,y)
    \right|
    \le \P\{A\} .
\end{align*}
Since the size of the dependency window is a compound Poisson process,
we can bound $\P\{A\}$ using the following lemma:

>>>>

First we quantify and prove the approximation \eqref{eqn:window_approx}.
The basic idea is that the probability is only in error if there is a chain of mutations
that create a dependency between the base of the T-mer and the flanking sequence.
To simplify the argument, first assume that all transition triples
change only a single, central site in the pattern,
taking into account context on each side of maximum length $w$.
(For instance, a trinucleotide model in which the flanking sites on each side
affect the mutation rate of the central site has $w=1$.)
%EM I suggest that we could re-introduce the mu^* notation again and remind people that these potential changes are uniform.
Recall that in our construction of the process we first chose the times and sites
at which potential changes might occur.
Suppose a potential change occurs at site $i-w$, at time $t$,
which therefore might result in a change in state at site $i$.
To resolve this, we need to know the outcome of any other potential changes
that might change the sequence between $i-w$ and $i + w$.
Consider those that might happen only to the right of $i$,
thus extending the sequence on which the change depends to the right.
The number of such potential changes is Poisson with mean $\mu^* w t$,
%EM Suggest "and their location is uniform..."
and will extend the amount of sequence on which the change at site $i$ depends
by a uniform random number of sites between 1 and $w$.
If one occurs, say at time $0 < s < t$,
then any further potential changes that occur between time 0 and time $s$
in the $w$ sites to the right of this older potential change
will again extend the range of dependency.
Putting this together, the amount of sequence to the right of a site
on which a mutation at that site depends
%EM Do we care that N is used for different things elsewhere?
is bounded by the compound Poisson process $\sum_{k=1}^N W_k$,
where $N$ is Poisson with mean $\mu^* w t$
and each $W_k$ is independent and uniform on $\{1, 2, \ldots, w\}$.

Now consider the probability of interest,
\begin{align*}
    \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \} .
\end{align*}
This probability can be written as a sum over all possible arrangements
and resolutions of potential changes.
Our approximation, $p_{n,\ell,r}(x,y)$ is the probability that would be obtained
by ignoring any potential changes that rely on context outside of the $n$-sequence $x$.
So, if $A$ is the event that the final subsequence $y$
depends on sequence outside of $x$ due to potential changes,
then
\begin{align*}
    \left|
        \P\{ X_{i+\ell}^{(n-\ell-r)}(t) = y \mid X_i^{(n)}(0) = x \}
        -
        p_{n,\ell,r}(x,y)
    \right|
    \le \P\{A\} .
\end{align*}
Since the size of the dependency window is a compound Poisson process,
we can bound $\P\{A\}$ using the following lemma:
